Semantic Preserving Bijective Mappings for Representations of Special Functions between Computer Algebra Systems and Document Preparation Systems

### 2018-11-15 main.tex
Give a short introduction to the concept of the translation in sec 4 (close #7)

\section{The Translator}
All translations are defined by a library (\gls*{csv} and \gls*{json} files) that defines translation patterns for each function and symbol. The pattern uses \verb|$i| as a placeholder to define the positions of the arguments. For example, the translation patterns for the Jacobi polynomial are illustrated in~\Cref{tab:placeholder_ex2}.

@@

These placeholders cause trouble when the \gls*{cas} uses the symbol \verb|$| for other reasons, e.g., the differentiation in \Maple{} is defined by
\begin{equation*}
\verb|diff(f, [x$n])|,
\end{equation*}
where $f$ is an algebraic expression or an equation, $x$ is the name of the differentiation variable, and $n$ is an integer representing the $n$-th order differentiation\footnote{\url{https://www.maplesoft.com/support/help/maple/view.aspx?path=diff}, seen 07/2018}. A translation for $\frac{d^2x^2}{dx^2}$ should look like this
\begin{equation*}
\verb|\deriv[2]{x^2}{x}| \overset{\langMaple}{\mapsto} \verb|diff(x^2, [x$2])|
\end{equation*}
but would end up as
\begin{equation*}
\verb|\deriv[2]{x^2}{x}| \overset{\langMaple}{\mapsto} \verb|diff(x^2, [xx])|.
\end{equation*}

@@

The \Macro s also allow one to specify optional arguments to distinguish between standard and another version of these functions. The Legendre and associated Legendre functions of the first kind are examples of such cases. The library that defines translations for each macro uses the macro name as the primary key to identify the translations. The Legendre and associated Legendre function of the first kind both use the same macro \verb|\LegendreP|. To distinguish such cases, we use a special syntax for such cases, shown in~\Cref{tab:legendreP-lex}.

The translator uses the \gls*{pom}-Tagger~\parencite{POM-Tagger}\footnote{Named according to the Part-of-Speech-Taggers in \gls*{nlp}.} to parse \LaTeX{} expressions into a parsed tree. The \gls*{pom}-Tagger is an \texttt{LL}-Parser defined by a context-free grammar in \gls*{bnf}. Each token will be tagged by meta information defined in lexicon files. We extend the lexicon files to provide also the information that is necessary for the translation process. An example of an entry of the lexicon file is given in~\Cref{tab:sine-lex-example}.

\begin{table}[ht!]
	\centering
	\begin{tabular}{lll}
	\hline
	\multicolumn{3}{l}{Symbol: \texttt{\textbackslash sin}} \\
	\! & \multicolumn{2}{l}{Feature Set: dlmf-macro} \\
	\! & \! & DLMF: \verb|\sin@@{z}|\\
	\! & \! & DLMF-Link: dlmf.nist.gov/4.14.E1\\
	\! & \! & Meanings: Sine\\
	\! & \! & Number of Parameters: 0\\
	\! & \! & Number of Variables: 1\\
	\! & \! & Number of Ats: 2\\
	\! & \! & Maple: \verb|sin($0)|\\
	\! & \! & Maple-Link: www.maplesoft.com/support/\\
	\! & \! & \hspace{32pt} help/maple/view.aspx?path=sin\\
	\! & \! & Mathematica: \verb|Sin[$0]|\\
	\! & \! & Mathematica-Link: reference.wolfram.com/\\
	\! & \! & \hspace{32pt} language/ref/Sin.html\\
	\hline
	\end{tabular}
	\caption{The entry of the trigonometric sine function in the lexicon file.}
	\label{tab:sine-lex-example}
\end{table} \section{Forward Translations}\label{sec:forward-translation}
An abstract translator class analyzes each node of the parsed tree and delegates them to specialized subtranslators. In this process, an object called \gls*{teo} is built. This \gls*{teo} is needed to rebuild a string representation of the mathematical expression after all nodes of the parsed tree were translated.

>>>>

\section{The Translator}
The translator analyzes a parse tree to perform translations. For generating a parse tree of \LaTeX{} expressions, the translator uses the \gls*{pom}-Tagger~\parencite{POM-Tagger}\footnote{Named according to the Part-of-Speech-Taggers in \gls*{nlp}.}. The \gls*{pom}-Tagger is an \texttt{LL}-Parser defined by a context-free grammar in \gls*{bnf}. Each token will be tagged by meta information defined in manually cultivated lexicon files. \gls*{cas} define their own syntax parser. We were able to use \Maple's internal data structure to get a parse tree of the input.

All translations are defined by a library (\gls*{csv} and \gls*{json} files) that defines translation patterns for each function and symbol. The pattern uses \verb|$i| as a placeholder to define the positions of the arguments. For example, the translation patterns for the Jacobi polynomial are illustrated in~\Cref{tab:placeholder_ex2}.

@@

The \Macro s also allow one to specify optional arguments to distinguish between standard and another version of these functions. The Legendre and associated Legendre functions of the first kind are examples of such cases. The library that defines translations for each macro uses the macro name as the primary key to identify the translations. The Legendre and associated Legendre function of the first kind both use the same macro \verb|\LegendreP|. To distinguish such cases, we use a special syntax, shown in~\Cref{tab:legendreP-lex}.

We integrated the defined translation patterns in the library also into the lexicon files of the \gls*{pom}-Tagger. Thus the tagger also tags a token by its translation patterns to other languages. \Cref{tab:sine-lex-example} gives an example of an entry of the lexicon file.

\begin{table}[ht!]
	\centering
	\begin{tabular}{lll}
	\hline
	\multicolumn{3}{l}{Symbol: \texttt{\textbackslash sin}} \\
	\! & \multicolumn{2}{l}{Feature Set: dlmf-macro} \\
	\! & \! & DLMF: \verb|\sin@@{z}|\\
	\! & \! & DLMF-Link: dlmf.nist.gov/4.14.E1\\
	\! & \! & Meanings: Sine\\
	\! & \! & Number of Parameters: 0\\
	\! & \! & Number of Variables: 1\\
	\! & \! & Number of Ats: 2\\
	\! & \! & Maple: \verb|sin($0)|\\
	\! & \! & Maple-Link: www.maplesoft.com/support/\\
	\! & \! & \hspace{32pt} help/maple/view.aspx?path=sin\\
	\! & \! & Mathematica: \verb|Sin[$0]|\\
	\! & \! & Mathematica-Link: reference.wolfram.com/\\
	\! & \! & \hspace{32pt} language/ref/Sin.html\\
	\hline
	\end{tabular}
	\caption{The entry of the trigonometric sine function in the lexicon file.}
	\label{tab:sine-lex-example}
\end{table}

\subsection{Escape the Placeholder Symbol}
The used placeholders cause trouble when the \gls*{cas} uses the symbol \verb|$| for other reasons, e.g., the differentiation in \Maple{} is defined by
\begin{equation*}
\verb|diff(f, [x$n])|,
\end{equation*}
where $f$ is an algebraic expression or an equation, $x$ is the name of the differentiation variable, and $n$ is an integer representing the $n$-th order differentiation\footnote{\url{https://www.maplesoft.com/support/help/maple/view.aspx?path=diff}, seen 07/2018}. A translation for $\frac{d^2x^2}{dx^2}$ should look like this
\begin{equation*}
\verb|\deriv[2]{x^2}{x}| \overset{\langMaple}{\mapsto} \verb|diff(x^2, [x$2])|
\end{equation*}
but would end up as
\begin{equation*}
\verb|\deriv[2]{x^2}{x}| \overset{\langMaple}{\mapsto} \verb|diff(x^2, [xx])|.
\end{equation*}

@@

\section{Forward Translations}\label{sec:forward-translation}
An abstract translator class analyzes each node of the parsed tree and delegates them to specialized subtranslators. In this process, an object called \gls*{teo} is built. This \gls*{teo} is needed to rebuild a string representation of the mathematical expression after all nodes of the parsed tree were translated.
